## Examples of  continuous distributions

### The exponential distribution

In Section \@ref(poisson), we considered the Poisson process, in which
events occur at random, at a rate $\theta$ per unit time.
The actual number of events which take place in any given
unit of time has $\text{Poisson}(\theta)$ distribution.
The *exponential* distribution represents
the time between consecutive events in this process.

Let $Y$ represent the time interval between two events.
Clearly this variable cannot be negative, but can take any
positive value.
The domain of $Y$ is $(0, \infty)$.
We have
\begin{align*}
P(Y > y) &= P(\text{no events in an interval of length $y$}) \\
&= \frac{e^{-\theta y} (\theta y)^0}{0!} \\
&= e^{-\theta y}
\end{align*}
so
\[F(y) = P(Y \leq y) = 1 - e^{-\theta y}, \quad y > 0.\]
Differentiating, we obtain the pdf
\[f(y) = \frac{d}{dy} F(y) = \theta e^{-\theta y}, \quad y > 0.\]

```{definition}
A random variable $Y$ has *exponential* distribution if it has
pdf of the form
\[f(y) = \theta e^{-\theta y}, \quad y > 0,\]
for some *rate parameter* $\theta > 0$.
We write $Y \sim \text{exponential}(\theta)$.
```

```{example}
 Suppose the lifetime in hours of a certain type of 
  electronic component is described by an Exponential
random variable with rate parameter $\theta = 0.01$. 
What is the probability such a component will
have a lifetime of between $100$ and $200$ hours?

The probability is the area under the curve $f(y) = 0.01 e^{-0.01 y}$ 
between $y = 100$ and $y = 200$, so
\begin{align*}
P(100 < Y \leq 200) &= \int_{100}^{200} 0.01e^{-0.01 y} dy \\
&= e^{-1} - e^{-2} = 0.37 - 0.14 = 0.23.
\end{align*}
```
We could find this in `R` with
```{r}
pexp(200, rate = 0.01) - pexp(100, rate = 0.01)
```
### The uniform distribution

The uniform distribution is one of the simplest probability 
distributions: it just places a constant density between
some range $(a, b)$:

```{definition}
A random variable $Y$ has *uniform* distribution if it has
pdf of the form
\[f(y) = \frac{1}{b-a}, a < y < b,\]
for some parameters $a, b \in \mathbb{R}$, with $b > a$.
We write $Y \sim U(a, b)$.
```

### The normal distribution

The normal distribution is probably the single most important distribution
in statistics. The main reason for its importance is the central limit
theorem, which you have seen before in MATH1024, and which
we will prove in Section \@ref(clt).

```{definition}
A random variable $Y$ has *normal* distribution if it has
pdf of the form
\[f(y) = \frac{1}{\sqrt{2 \pi \sigma^2}}
\exp\left\{-\frac{1}{2 \sigma^2} (y - \mu)^2 \right\},\]
for some parameters $\mu \in \mathbb{R}$ and $\sigma^2 > 0$.
We write $Y \sim N(\mu, \sigma^2)$.
```

```{example}
Suppose that daily water use at a factory varies about a mean of 
$77500$ litres with standard
deviation $5700$ litres. If demand is normally distributed

1. What proportion of days does the demand fall short of $70000$ litres?
1. What proportion of days does demand exceed $90000$ litres?
1. What is your reaction to a demand of $175000$ gallons?

Writing $X$ for the daily water use in litres
, we have $X \sim N(77500, 5700^2)$
We can use `R` to compute the probability $P(X < 70000) = F(70000)$
```{r}
pnorm(70000, mean = 77500, sd = 5700)
```
so the daily water use will be less than $70000$ litres
about $9.4\%$ of the time.

We can find $P(x > 90000) = 1 - F(90000)$
```{r}
1 - pnorm(90000, mean = 77500, sd = 5700)
```
so the daily water use will be more than $90000$ litres
about $1.4\%$ of the time.


We can find  $P(x > 175000) = 1 - F(175000)$

```{r}
1 - pnorm(175000, mean = 77500, sd = 5700)
```
In fact, the number is not exactly zero, but it is so
small that the computer is rounding it to zero.
Such an extreme water use is therefore surprising,
and an
explanation should be sought. It is possible that a
 error has occurred in recording the water use, such
as two days data being taken together. Alternatively, perhaps
our model which assumes that 
$X \sim N(77500, 5700^2)$ is incorrect.
This idea of surprise at an extreme result of
low probability, as predicted by a statistical model, 
will be important later in this
module and also in modules such as MATH2010 Statistical Modelling I.

