# Moments

## Expected value

Suppose we have a discrete random variable $Y$ with probability function 
$p(.)$ with domain $D$.
Although $p(.)$ tells us everything about the properties of $Y$, 
it is often useful to
summarise the properties $Y$ using a few simple quantities.

A simple summary of the probabilistic properties of $Y$ is the 
*expected value* or
(population) *mean* of $Y$, denoted by $E(Y)$ or $\mu$,
depending on the context.

```{definition}
If $Y$ is a discrete random variable with probability function $p(.)$
and domain $D$, the expected value of $Y$ is
\[\mu = E(Y) = \sum_{y \in D} y p(y).\]
If $Y$ is a continuous random variable with probability density 
function $f(.)$ and domain $D$, 
 the expected value of Y is:
\[\mu = E(Y) = \int_{-\infty}^\infty yf(y)dy = \int_D y f(y) dy.\]
```

```{example, bernoullimean, name = "Expected value of the Bernoulli distribution"}
Suppose $Y \sim \text{Bernoulli}(\theta)$. Then
\[E(Y) = \sum_{y \in \{0, 1\}} y p(y) = 0 \times (1- \theta) + 1 \times \theta = \theta.\]
```

```{example, expmean, name = "Expected value of the exponential distribution"}
Suppose $Y \sim \text{exponential}(\theta)$. Then
\begin{align*}
E(Y) &= \int_{0}^\infty y \theta e^{-\theta y} dy  \\
&= \frac{1}{\theta} \int_{0}^\infty t e^{-t} dt \\
&= \frac{1}{\theta} \left\{[-t e^{-t}]_0^\infty + \int_0^\infty e^{-t}dt \right\} \\
&= \frac{1}{\theta} \{0 + 1\} = \frac{1}{\theta}.
\end{align*}
Note that we have shown $\int_{0}^\infty t e^{-t} dt  = 1$: this result will
be useful later on.
```

We can find the expected value of new random variable $h(Y)$ with
\[E[h(Y)] = \int_{-\infty}^{\infty} h(y) f(y) dy = \int_D h(y) f(y) dy\]
  if $Y$ is a continous random variable with pdf $f(.)$. We may
  obtain a similar expression in the discrete case.


## Variance
  
```{definition}
The (population) variance of $Y$ is
\[ \text{Var}(Y) = E\left\{[Y - E(Y)]^2 \right\}.\]
```

It is easy to show that
\[ \text{Var}(Y) = E(Y^2) - [E(Y)]^2.\]
$\text{Var}(Y)$ is a measure of spread, and is often denoted by $\sigma^2$.
  
We sometimes use the (population) standard
deviation, which is just the square root of the variance, to return to the original
scale of measurement of $Y$.

```{example, name = "Variance of the Bernoulli distribution"}
Suppose $Y \sim \text{Bernoulli}(\theta)$. We have seen in Example
\@ref(exm:bernoullimean) that $E(Y) = \theta$. We have
\[E(Y^2) = \sum_{y \in \{0, 1\}} y^2 p(y) = 0^2 \times (1- \theta) + 1^2 \times \theta
= \theta,\]
so
\[\text{Var}(Y) = E(Y^2) - [E(Y)]^2 = \theta - \theta^2 = \theta ( 1 - \theta).\]
```

```{example, expvar, name = "Variance of the exponential distribution"}
Suppose $Y \sim \text{exponential}(\theta)$. We have seen in Example
\@ref(exm:expmean) that $E(Y) = 1/\theta$. We have
\begin{align*}
E(Y^2) &= \int_{0}^\infty y^2 \theta e^{-\theta y} dy  \\
&= \frac{1}{\theta^2} \int_{0}^\infty t^2 e^{-t} dt \\
&= \frac{1}{\theta} \left\{[-t^2 e^{-t}]_0^\infty + 2 \int_0^\infty t e^{-t}dt \right\} \\
&= \frac{1}{\theta^2} \{0 + 2\} = \frac{2}{\theta^2},
\end{align*}
since we saw in Example \@ref(exm:expmean) that
$\int_0^\infty t e^{-t}dt = 1$.
So
\[\text{Var}(Y) = E(Y^2) - [E(Y)]^2 = \frac{2}{\theta^2} - \frac{1}{\theta^2} 
= \frac{1}{\theta^2}.\]
```

## Higher-order moments

So far we have summarised random variables using the mean and variance (or standard
deviation), which measure the “location” and the “spread”.
Why would we need anything more? 
Consider the following two density functions of two different continuous
distributions:

```{r, echo = FALSE}
par(mfrow = c(1, 2))
curve(dnorm(x), from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
curve(dunif(x, min = -sqrt(3), max = sqrt(3)), from = -6, to = 6,
      xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
```

In fact, the plot on the left is the pdf of a $N(0, 1)$ distribution,
and the plot on the right is the pdf of a $\text{Uniform}(-\sqrt{3}, \sqrt{3})$
distribution. Both have mean $0$ and variance $1$, yet they look very different.
They are both symmetric about $0$ (the mean) but differ in terms of "shape".

Consider two more density functions:

```{r, echo = FALSE}
par(mfrow = c(1, 2))
curve(dgamma(x + 2, shape = 4, rate = 2),
      from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
b <- sqrt(7)/4
curve(0.75 * dnorm(x, mean = 0.75, sd = b) + 0.25 * dnorm(x, mean = -2.25, sd = b),
      from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
```

Again both have mean $0$ and variance $1$, yet they look very different.
Neither is symmetric and they have different “shapes”.

How can we capture something useful about “shape”?
We use so-called *higher-order moments* --- particularly the third and fourth moments.
So we need a general definition of moments and it will be useful to obtain
relationships between them.

In what follows, we will assume our random variable $Y$ has continuous distribution. 
The discrete case
follows by replacing the pdf by the probability function and the integral by a
sum.

```{definition}
The *$r$th moment about the origin* is
\[\mu_r^\prime = E(Y^r) = \int_{-\infty}^\infty y^r f(y) dy\]
and the 
*$r$th moment about the mean* is
\[\mu_r = E\left\{[Y – E(Y)]^r \right\} =\int_{-\infty}^\infty (y - \mu)^r f(y) dy.\]
```

We have
\begin{align*}
\mu_1^\prime &= \mu = E(Y) \\
\mu_1 &= 0 \\
\mu_2 &= \text{Var}(Y).
\end{align*}

How about the third and fourth moments about the mean?
We have
\[\mu_3 = E\left\{[Y - E(Y)]^3\right\} = \int_{-\infty}^\infty (y - \mu)^3 f(y) dy.\]

```{theorem, mu3}
We have
\[\mu_3 = \mu_3^\prime - 3 \mu_2^\prime \mu + 2 \mu^3.\]
```
```{remark}
This formula allows us to find the third moment about the mean from the first
three moments about the origin.
````
```{proof}
We have
\[\mu_3 = E\{[Y - E(Y]^3\} = E\{(Y - \mu)^3\},\]
writing $\mu = E(Y)$. Expanding, we have
\[(Y - \mu)^3 = Y^3 - 3 Y^2 \mu + 3 Y \mu^2  - \mu^3\]
and using the linearity of expectation gives
\begin{align*}
\mu_3 &= E(Y^3) - 3 \mu E(Y^2) + 3 \mu^2 E(Y) - \mu^3 \\
&= \mu_3^\prime - 3 \mu \mu_2^\prime + 3 \mu^2 \mu - \mu^3 \\
&= \mu_3^\prime - 3 \mu \mu_2^\prime + 2 \mu^3,
\end{align*}
as required.
```

If $Y$ has symmetric distribution then $\mu_3 = 0$.
If $Y$ has a heavier right tail than left tail then $\mu_3 > 0$,
and conversely if $Y$ has a heavier left tail than right, then $\mu_3 < 0$.

Similarly, the fourth moment about the mean is
\[\mu_4 =  E\left\{[Y - E(Y)]^4\right\} = \int_{-\infty}^\infty (y - \mu)^4 f(y) dy.\]
```{theorem, mu4}
We have
\[\mu_4 = \mu_4^\prime - 4 \mu_3^\prime  \mu + 6 \mu_2^\prime \mu^2 - 3 \mu^4.\]
```
```{remark}
This formula allows us to find the fourth moment about the mean from the first
fourth moments about the origin.
```
The proof is very similar to that of \@ref(thm:mu3), and we
leave it as an exercise.

For symmetric distributions, roughly speaking
thick tails lead to higher values of $\mu_4$ than light tails.

## Standardised moments

Remember that the basic idea is to describe location and spread via the mean and
variance and the describe "shape" in terms of the third and fourth moments.
So we don't mix up spread with shape we usually use standardised third and fourth
moments about the mean.

```{definition}
The *skewness* is 
\[\gamma_1 = \frac{\mu_3}{\mu_2^{3/2}},\]
the standardised third moment about the mean.
```
```{definition}
The *kurtosis* is
\[\gamma_2 = \frac{\mu_4}{\mu_2^2}\]
the standardised fourth moment about the mean
```

The skewness and kurtosis are unchanged by linear transformations of $Y$.

Of course we could consider yet higher order moments to fine tune our
understanding of $Y$.
However, we often stop at $r = 4$.
Even so, if we just want to obtain the first four moments of a distribution, 
this may involve a
lot of (difficult!) integration.
In Chapter \@ref(genfuns) we will find a method that allows us to find as many moments as
we like but with only one integration required.

We will see in Example \@ref(exm:cgfnormal) that
any $N(\mu, \sigma^2)$ distribution has skewness $\gamma_1 = 0$ and kurtosis
$\gamma_2 = 3$. The kurtosis of other distributions is often compared with
that of a normal distribution: if $\gamma_2 < 3$, a distribution has 
lighter tails than the normal distribution, while if $\gamma_2 > 3$, a distribution
has heavier tails than the normal distribution.

```{example}
Let $Y \sim \text{Bernoulli}(\theta)$.
It is easy to see that $\mu_r^\prime = \theta$, for 
$r = 1, 2, 3, \ldots$.

So, using Theorem \@ref(thm:mu3), the third moment about the mean is
\begin{align*}
\mu_3 &= \mu_3^\prime - 3 \mu_2^\prime \mu + 2 \mu^3 \\
&= \theta - 3 \theta . \theta + 2 \theta^3 \\
&= \theta - 3 \theta^2 + 2\theta^3 \\
&= \theta (1 - \theta) (1 - 2 \theta).
\end{align*}
So the skewness is 
\[\gamma_1 = \frac{\mu_3}{\mu_2^{3/2}} 
= \frac{\theta (1 - \theta)(1 - 2 \theta)}{[\theta (1 - \theta)]^{3/2}}
= \frac{1 - 2 \theta}{\sqrt{\theta(1 - \theta)}}.\]
Note that the skewness is positive if $\theta < 0.5$, zero if $\theta = 0.5$,
and negative if $\theta > 0.5$.

Using Theorem \@ref(thm:mu4), the fourth moment about the mean is
\begin{align*}
\mu_4 &= \mu_4^\prime - 4 \mu_3^\prime \mu + 6 \mu_2^\prime \mu^2 - 3 \mu^4 \\
&= \theta - 4 \theta. \theta + 6 \theta. \theta^2 - 3 \theta^4 \\
&= \theta(1 - 4 \theta + 6 \theta^2 - 3 \theta^3) \\
&= \theta (1 - \theta) (1 - 3 \theta + 3 \theta^2).
\end{align*}
So the kurtosis is
\[\gamma_2 = \frac{\mu_4}{\mu_2^2} = \frac{1 - 3 \theta + 3 \theta^2}{\theta (1 - \theta)}.\]
Note that $\gamma_2 = 1$ for $\theta = 0.5$.
```


```{example, homexp}

Let $Y \sim \text{exponential}(\theta)$, with pdf 
$f (y) = \theta \exp(- \theta y)$ for $y > 0$.
We have seen that $E(Y) = \theta^{-1}$ (Example \@ref(exm:expmean))
and $E(Y^2) = 2 \theta^{-2}$ (Example \@ref(exm:expvar)).

The third moment is
\begin{align*}
\mu_3^\prime = E(Y^3) &= \int_{0}^\infty y^3 \theta e^{-\theta y} dy \\
&= \frac{1}{\theta^2} \int_0^\infty t^3 e^{-t} dt \\
&= \frac{1}{\theta^3} \left\{\left[-t^3 e^{-t}\right]_0^\infty
+ 3 \int_0^\infty t^2 e^{-t} dt \right\} \\
&= \frac{1}{\theta^3} \left\{0 + 3 \times 2 \right\} = \frac{6}{\theta^3},
\end{align*}
where we have used the fact that $\int_0^\infty t^2 e^{-t} dt = 2$, 
from Example \@ref(exm:expvar).
So, using Theorem \@ref(thm:mu3), the third moment about the mean is
\[ \mu_3 = \mu_3^\prime - 3\mu_2^\prime \mu + 2 \mu^3
= \frac{6}{\theta^3} - \frac{6}{\theta^2} \cdot \frac{1}{\theta} + \frac{2}{\theta^3}
= \frac{2}{\theta^3}.\]
So the skewness is
\[\gamma_1 = \frac{(2/ \theta^3)}{(1/ \theta^2)^{3/2}}
 = 2,\]
 so the exponential distribution has the same positive skewness for all
 values of the rate parameter $\theta$.
 
 The fourth moment is
 \begin{align*}
 \mu_4^\prime = E(Y^4) &= \int_0^\infty y^4 \theta e^{-\theta y} dy \\
 &= \frac{1}{\theta^4} \int_0^\infty t^4 e^{-t} dt \\
 &= \frac{1}{\theta^4} \left\{[-t^4 e^{-t}]_0^\infty + 
 4 \int_0^\infty t^3 e^{-t} dt \right\} \\
 &= \frac{1}{\theta^4} \left\{0 + 4 \times 6 \right\}
 = \frac{24}{\theta^4},
 \end{align*}
where we have used that $\int_0^\infty t^3 e^{-t}dt = 6$, from above.
So, using Theorem \@ref(thm:mu4),
the fourth moment about the mean is
\begin{align*}
\mu_4 &= \mu_4^\prime - 4 \mu_3^\prime \mu + 6 \mu_2^\prime \mu^2 - 3 \mu^4\\
&= \frac{24}{\theta^4} - 4 \cdot \frac{6}{\theta^3} \cdot \frac{1}{\theta}
+ 6 \cdot \frac{2}{\theta^2} \cdot \frac{1}{\theta^2}
- \frac{3}{\theta^4} \\
&= \frac{9}{\theta^4}.
\end{align*}
 So the kurtosis is
 \[\gamma_2 = \frac{9/\theta^4}{(1/\theta^2)^2} =  9,\]
 which does not depend on $\theta$. All exponential random variables
 are positively skewed ($\gamma_1 = 2$), with a high
 kurtosis ($\gamma_2 = 4$), meaning the exponential distribution
 has heavier tails than the normal distribution.
```
