# (PART) Properties of distributions {-}

# Moments

## Expected value

Suppose we have a discrete rv $Y$ with probability function 
$p(.)$ with domain $D$.
Although $p(.)$ tells us everything about the properties of $Y$, 
it is often useful to
summarise the properties $Y$ using a few simple quantities.
This is similar to the situation in applications where we collect
a large amount of
data and we use summary statistics (e.g. the sample mean, 
the sample standard
deviation) to help us to understand or interpret the data.

A simple summary of the probabilistic properties of $Y$ is the 
*expected value* or
(population) *mean* of $Y$, denoted by $E(Y)$ or $\mu$,
depending on the context.

If we think of probability as mass distributed at various 
discrete points on a
numerical scale, then $E(Y)$ is simply the centre of mass of $p(.)$.
\[E(Y) = \sum_{y \in D} y p(y).\]

**TODO: add explanation in terms of sample mean**

```{example, name = "Expected value of the Bernoulli distribution"}
```

If $Y$ is a continuous rv with probability density 
function $f(.)$ and domain $D$, 
 the expected value of Y is:
\[E(Y) = \int_{-\infty}^\infty yf(y)dy = \int_D y f(y) dy.\]


```{example, name = "Expected value of the exponential distribution"}
```

We can find the expected value of new random variable $h(Y)$ with
\[E[h(Y)] = \int_{-\infty}^{\infty} h(y) f(y) dy = \int_D h(y) f(y) dy\]
  if $Y$ is a continous random variable with pdf $f(.)$. We may
  obtain a similar expression in the discrete case.


## Variance
  
We can define the (population) variance of $Y$ by
\[ \text{Var}(Y) = E\left\{[Y - E(Y)]^2 \right\}.\]
It is easy to show that
\[ \text{Var}(Y) = E(Y^2) - [E(Y)]^2.\]
$\text{Var}(Y)$ is a measure of spread. It is often denoted by $\sigma^2$.
  
We sometimes use the (population) standard
deviation, which is just the square root of the variance, to return to the original
scale of measurement of $Y$.

```{example, name = "Variance of the Bernoulli distribution"}
```

```{example, name = "Variance of the exponential distribution"}
```

## Higher-order moments

So far we have summarised rvs using the mean and variance (or standard
deviation), which measure the “location” and the “spread”.
Why would we need anything more? 
Consider the following two density functions of two different continuous
distributions:

```{r, echo = FALSE}
par(mfrow = c(1, 2))
curve(dnorm(x), from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
curve(dunif(x, min = -sqrt(3), max = sqrt(3)), from = -6, to = 6,
      xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
```

In fact, the plot on the left is the pdf of a $N(0, 1)$ distribution,
and the plot on the right is the pdf of a $\text{Uniform}(-\sqrt{3}, \sqrt{3})$
distribution. Both have mean $0$ and variance $1$, yet they look very different.
They are both symmetric about $0$ (the mean) but differ in terms of “shape”.

Consider two more density functions:

```{r, echo = FALSE}
par(mfrow = c(1, 2))
curve(dgamma(x + 2, shape = 4, rate = 2),
      from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
b <- sqrt(7)/4
curve(0.75 * dnorm(x, mean = 0.75, sd = b) + 0.25 * dnorm(x, mean = -2.25, sd = b),
      from = -6, to = 6, xlab = "x", ylab = "f(x)", ylim = c(0, 0.5))
```

Again both have mean $0$ and variance $1$, yet they look very different.
Neither is symmetric and they have different “shapes”.

How can we capture something useful about “shape”?
We use so-called *higher-order moments* --- particularly the third and fourth moments.
So we need a general definition of moments and it will be useful to obtain
relationships between them.

In what follows, we will assume our rv $Y$ has continuous distribution. 
The discrete case
follows by replacing the pdf by the probability function and the integral by a
sum.

In general, we define the
*$r$th moment about the origin*
\[\mu_r^\prime = E(Y^r) = \int_{-\infty}^\infty y^r f(y) dy\]
and the 
*$r$th moment about the mean*
\[\mu_r = E\left\{[Y – E(Y)]^r \right\} =\int_{-\infty}^\infty (y - \mu)^r f(y) dy.\]

We have
\begin{align*}
\mu_1^\prime &= \mu = E(Y) \\
\mu_1 &= 0 \\
\mu_2 = \text{Var}(Y).
\end{align*}

How about the third and fourth moments about the mean?
We have
\[\mu_3 = E\left\{[Y - E(Y)]^3\right\} = \int_{-\infty}^\infty (y - \mu)^3 f(y) dy.\]

```{theorem}
We have
\[\mu_3 = \mu_3^\prime - 3 \mu_2^\prime \mu + 2 \mu^3.\]
```
```{remark}
This formula allows us to find the third moment about the mean from the first
three moments about the origin.
````
```{proof}
**TODO: add proof of this (or gap for this?)**
```

Similarly
\[\mu_4 =  E\left\{[Y - E(Y)]^4\right\} = \int_{-\infty}^\infty (y - \mu)^4 f(y) dy.\]
```{theorem}
We have
\[\mu_4 = \mu_4^\prime - 4 \mu_3^\prime  \mu + 6 \mu_2^\prime \mu^2 - 3 \mu^4.\]
```
```{remark}
This formula allows us to find the fourth moment about the mean from the first
fourth moments about the origin.
```

If $Y$ has symmetric distribution then $\mu_3 = 0$.
If $Y$ has a heavier right tail than left tail then $\mu_3 > 0$,
and conversely if $Y$ has a heavier left tail than right, then $\mu_3 < 0$.

For symmetric distributions, roughly speaking
thick tails lead to higher values of $\mu_4$ than light tails.

## Standardised moments

Remember that the basic idea is to describe location and spread via the mean and
variance and the describe “shape” in terms of the third and fourth moments.
So we don’t mix up spread with shape we usually use standardised third and fourth
moments about the mean.

**TODO: could add some examples of what goes wrong if we just use third
and fourth moments about the mean directly**

The **skewness** is the standardised third moment about the mean, defined
as
\[\gamma_1 = \frac{\mu_3}{\mu_2^{3/2}}.\]
The **kurtosis** is the standardised fourth moment about the mean, defined
as
\[\gamma_2 = \frac{\mu_4}{\mu_2^2}.\]
Is it obvious that these quantities are unchanged by linear transformations of $Y$?

Of course we could consider yet higher order moments to fine tune our
understanding of $Y$.
However, we often stop at $r = 4$.
Even so, if we just want to obtain the first four moments of a distribution, 
that may involve a
lot of (difficult!) integration.
In Chapter **TODO ????** we will find a method that allows us to find as many moments as
we like but with only one integration required.

```{example, name = "Skewness and kurtosis of the Bernoulli distribution"}

Let $Y \sim \text{Bernoulli}(\theta)$.
It is easy to see that $\mu_r^\prime = \theta$.

So we get:
**TODO: finish/leave gap**
```

```{example, name = "Skewness and kurtosis of the Exponential distribution"}

Let $Y \sim \text{Exponential}(\theta)$, with pdf 
$f (y) = \theta exp(- \theta y)$ for $y > 0$.
We have seen **TODO ???** that $E(Y) = \theta^{-1}$
and $E(Y^2) = 2 \theta^{-2}$.

So we get:
**TODO: finish/leave gap**.
```
